---
title: "Visualizations of COVID-19 cases in Africa"
author: "Julius Okwir"
date: "2020/05/06"
output:
    html_document:
         code_folding: hide
---

## Background
Late in December 2019, the [World Health Organisation](https://www.who.int/emergencies/diseases/novel-coronavirus-2019/events-as-they-happen) (WHO) China Country Office obtained information about a severe pneumonia of an unknown cause, detected in the city of Wuhan in Hubei province, China. This later turned out to be the **novel coronavirus disease** (COVID-19), an infectious disease caused by **severe acute respiratory syndrome coronavirus-2** (SARS-CoV-2) of the corona virus family. The disease causes respiratory illness characterised by primary [symptoms](https://www.cdc.gov/coronavirus/2019-ncov/symptoms-testing/symptoms.html) like cough, fever, and in more acute cases, difficulty in breathing. **WHO** later [declared covid-19 as a Pandemic](https://time.com/5791661/who-coronavirus-pandemic-declaration/) because of its fast rate of spread across the Globe with over [3.58 Million confirmed cases and over 250,000 deaths](https://coronavirus.jhu.edu/map.html) as of May 05, 2020. The African continent started confirming its first cases in late January and mid-February of 2020 in some countries. The disease has since spread across 53 of the 54 Africa countries with over [46,000 confirmed cases and over 1,800 deaths](https://africanarguments.org/2020/04/07/coronavirus-in-africa-tracker-how-many-cases-and-where-latest/) as of May 05, 2020. 

## The Dataset
The [covid_19_africa.csv](https://github.com/ju-ok/Covid-19-Africa.git) dataset contains daily level information about the Covid-19 cases in Africa.
It is a time series data and hence the number of cases on any given day is cumulative.I extracted the data from the [covid_19_data.csv](https://www.kaggle.com/sudalairajkumar/novel-corona-virus-2019-dataset#covid_19_data.csv) which was made available on kaggle. The R script that I used to prepare this dataset is also available on [my Github repository](https://github.com/ju-ok/Covid-19-Africa.git). The original datasets can be found on [John Hopkins University Github repository](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_daily_reports) for covid_19. 

### Field description
- ObservationDate: YY/MM/DD
- Country: name of African country
- Region: Region of the African country
- Confirmed: the number of cOVID-19 confirmed cases
- Deaths:  the number of deaths from COVID-19
- Recovered: the number of recovered cases
- Active: the number of people still infected with COVID-19

## Task
**Visualizations of Africa’s COVID-19 confirmed cases, deaths, and recovered cases by country.** 

## Data processing
**Loading the required packages and setting a general theme for all the plots.**
```{r, message = FALSE, Warning = FALSE}
suppressMessages(library(tidyverse))
suppressMessages(library(lubridate))
suppressMessages(library(DT))
suppressMessages(library(plotly))
theme_set(theme_classic())
```

**Importing the dataset**
```{r covid_19_africa, echo = TRUE, message = FALSE}
covid_africa <- read_csv("covid_19_africa.csv")

# converting the classes of Confirmed, Deaths, and Recovered columns to class integer
covid_africa <- covid_africa %>%
        
        mutate(Region = as_factor(Region),
               Confirmed = as.integer(Confirmed), 
               Deaths = as.integer(Deaths),
               Recovered = as.integer(Recovered),
               Active = as.integer(Active))
```

## COVID-19 cases in Africa by Country from 27th of January to date
```{r covid-19 Africa}
datatable(covid_africa)
```

## Number of African countries with confirmed cases
```{r confirmed africa, echo = TRUE}
length(unique(covid_africa$Country))
```
At the time of updating this script, only 52 of the 54 African countries recognized by the UN and WHO had confirmed cases.

## Latest number of cases of COVID-19 in Africa by country 
Note that the latest number of cases on a given day is the cumulative sum of cases reported by end of the previous day.
```{r latest cases, echo = TRUE}
latest <- covid_africa %>%
        filter(ObservationDate == today() - 1 )
datatable(latest)
```

## VISUALIZATIONS OF COVID-19 CASES IN AFRICA
**Interactive plot of weekly changes in the number COVID-19 cases in Africa**
```{r number of cases in africa, echo = TRUE, fig.width=6, fig.height=8}
weekly_cases <- covid_africa %>%
        # grouping by observationdate
        group_by(ObservationDate) %>%
        
        # obtaining the sum of cases for each day
        summarise(Confirmed = sum(Confirmed), Deaths = sum(Deaths), Recovered = sum(Recovered), Active = sum(Active))

# creating a character vector of colors for the cases
colors <- c("Confirmed" = "orange", "Deaths" = "red", "Recovered" = "green", "Active" = "cyan2")

# plotting
p <- weekly_cases %>%
        
        # placing all cases in a single column and their sums in another column 
        gather(Confirmed, Deaths, Recovered, Active, key = "cases", value = "number") %>%
        
        # specifying the values on the x and y axes
        ggplot(aes(x = ObservationDate, y = number)) +
        
        # specifying the plot type
        geom_line(aes(color = cases)) +
        
        # setting the date labels
        scale_x_date(date_labels = "%d-%b", date_breaks = "1 week") +
        
        # manually changing the colors of the cases
        scale_color_manual(values = colors) + 
        
        # setting parameters for x axis, plot title, and plot margin
        theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0, size = 10, face = "bold"),
        plot.margin = margin(1, 0, 0, 0, unit = "cm")) +
        
        # setting the plot labels
        labs(title = "Weekly changes in the number of COVID-19 cases in Africa",
             x = "Date", y = "Number of cases")

# making the plot interactive
ggplotly(p)
```

**Confirmed number of cases by country.**
```{r confirmed by country, echo = TRUE, fig.width=10, fig.height=12}
(latest_confirmed <- latest %>%
                
                # reordering the countries by number of confirmed cases i.e highest to lowest
                mutate(Country = fct_reorder(Country, Confirmed)) %>%
                
                # specifying the values on the x and y axes 
                ggplot(aes(x = Country, y = Confirmed)) +
                
                # setting parameters for the bar plots of confirmed cases by country
                geom_bar(stat = "identity", show.legend = FALSE, fill = "orange", width = 0.6) + 
                
                # specifying and positioning of the bar labels
                geom_text(aes(label = Confirmed), hjust =  0, size = 3) +
                
                # flipping the x and y co-ordinates
                coord_flip(clip = "off", expand = FALSE) +
                
                # specifying desired parameters for the plot title and margin
                theme(plot.title = element_text(hjust = 0, size = 10, face = "bold"),
                      plot.margin = margin(0, 1, 0, 0, unit = "cm")) +
                
                # specifying the plot title
                labs(title = "Africa's Confirmed COVID-19 cases by Country. 2020/05/06"))
```

**Confirmed deaths by country.**
```{r confirmed deaths, echo = TRUE, fig.width=10, fig.height=12}
(latest_deaths <- latest %>%
                
                # reordering the countries by number of confirmed deaths i.e highest to lowest
                mutate(Country = fct_reorder(Country, Deaths)) %>%
                
                # specifying the values on the x and y axes 
                ggplot(aes(x = Country, y = Deaths)) +
                
                # setting parameters for the bar plots of confirmed deaths by country
                geom_bar(stat = "identity", show.legend = FALSE, fill = "red", width = 0.6) +
                
                # specifying and positioning of the bar labels
                geom_text(aes(label = Deaths), hjust =  0, size = 3) +
                
                # flipping the x and y co-ordinates
                coord_flip(clip = "off", expand = FALSE) +
                
                # specifying desired parameters for the plot title and margin
                theme(plot.title = element_text(hjust = 0, size = 10, face = "bold"),
                      plot.margin = margin(0, 1, 0, 0, unit = "cm")) +
                
                # specifying the plot title
                labs(title = "Africa's Deaths from COVID-19 by Country. 2020/05/06"))

```

**Recovered cases by country.**
```{r recovered, echo = TRUE, fig.width=10, fig.height=12}
(latest_recovered <- latest %>%
                
                # reordering the countries by number of recovered cases i.e highest to lowest
                mutate(Country = fct_reorder(Country, Recovered)) %>%
                
                #  specifying the values on the x and y axes 
                ggplot(aes(x = Country, y = Recovered)) +
                
                # setting parameters for the bar plots of confirmed recoveries by country
                geom_bar(stat = "identity", show.legend = FALSE, fill = "green", width = 0.6) + 
                
                # specifying and positioning of the bar labels
                geom_text(aes(label = Recovered), hjust =  0, size = 3) +
                
                # flipping the x and y  co-ordinates
                coord_flip(clip = "off", expand = FALSE) +
                
                # specifying desired parameters for the plot title and margin
                theme(plot.title = element_text(hjust = 0, size = 10, face = "bold"),
                      plot.margin = margin(0, 1, 0, 0, unit = "cm")) +
                
                # specifying the plot title
                labs(title = "Africa's Recovered cases from COVID-19 by Country. 2020/05/06"))
```

**Active cases by country.**
```{r active, echo = TRUE, fig.width=10, fig.height=12}
(latest_active <- latest %>%
                
                # reordering the countries by number of active cases i.e highest to lowest
                mutate(Country = fct_reorder(Country, Active)) %>%
                
                #  specifying the values on the x and y axes 
                ggplot(aes(x = Country, y = Active)) +
                
                # setting parameters for the bar plots of active cases by country
                geom_bar(stat = "identity", show.legend = FALSE, fill = "cyan2", width = 0.6) + 
                
                # specifying and positioning of the bar labels
                geom_text(aes(label = Active), hjust =  0, size = 3) +
                
                # flipping the x and y  co-ordinates
                coord_flip(clip = "off", expand = FALSE) +
                
                # specifying desired parameters for the plot title and margin
                theme(plot.title = element_text(hjust = 0, size = 10, face = "bold"),
                      plot.margin = margin(0, 1, 0, 0, unit = "cm")) +
                
                # specifying the plot title
                labs(title = "Africa's Active cases of COVID-19 by Country. 2020/05/06"))
```

## References
1. John Hopkins University Covid_19 datasets: 
https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_daily_reports
2. SRk's dataset on kaggle: https://www.kaggle.com/sudalairajkumar/novel-corona-virus-2019-dataset#covid_19_data.csv
3. African Arguments, Covid-19 Africa Tracker: 
https://africanarguments.org/2020/04/07/coronavirus-in-africa-tracker-how-many-cases-and-where-latest/











