---
title: "EDA: Visualizations of COVID-19 cases in Africa"
author: "Julius Okwir"
date: "2020/05/20"
output:
    html_document:
         code_folding: hide
---

## Background
Late in December 2019, the [World Health Organisation](https://www.who.int/emergencies/diseases/novel-coronavirus-2019/events-as-they-happen) (WHO) China Country Office obtained information about a severe pneumonia of an unknown cause, detected in the city of Wuhan in Hubei province, China. This later turned out to be the **novel coronavirus disease** (COVID-19), an infectious disease caused by **severe acute respiratory syndrome coronavirus-2** (SARS-CoV-2) of the corona virus family. The disease causes respiratory illness characterised by primary [symptoms](https://www.cdc.gov/coronavirus/2019-ncov/symptoms-testing/symptoms.html) like cough, fever, and in more acute cases, difficulty in breathing. **WHO** later [declared covid-19 as a Pandemic](https://time.com/5791661/who-coronavirus-pandemic-declaration/) because of its fast rate of spread across the Globe with over [4.9 Million confirmed cases and over 320,000 deaths](https://coronavirus.jhu.edu/map.html) as of May 20, 2020. The African continent started confirming its first cases in late January and mid-February of 2020 in some countries. The disease has since spread across all the 54 Africa countries with over [91,000 confirmed cases and over 2,900 deaths](https://africanarguments.org/2020/04/07/coronavirus-in-africa-tracker-how-many-cases-and-where-latest/) as of May 20, 2020. 

## The Dataset
The [covid_19_africa.csv](https://github.com/ju-ok/Covid-19-Africa.git) dataset contains daily level information about the Covid-19 cases in Africa.
It is a time series data and hence the number of cases on any given day is cumulative.I extracted the data from the [covid_19_data.csv](https://www.kaggle.com/sudalairajkumar/novel-corona-virus-2019-dataset#covid_19_data.csv) which was made available on kaggle. The R script that I used to prepare this dataset is also available on [my Github repository](https://github.com/ju-ok/Covid-19-Africa.git). The original datasets can be found on [John Hopkins University Github repository](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_daily_reports) for covid_19. 

### Field description
- ObservationDate: YY/MM/DD
- Country: name of African country
- Region: Region of the African country
- Confirmed: the number of cOVID-19 confirmed cases
- Deaths:  the number of deaths from COVID-19
- Recovered: the number of recovered cases
- Active: the number of people still infected with COVID-19

## Task
**Visualizations of Africa’s COVID-19 confirmed cases, deaths, recovered, and active  cases.** 

## Data processing
```{r, message = FALSE, Warning = FALSE, message=FALSE}
# loading the required packages
suppressMessages(library(tidyverse))
suppressMessages(library(lubridate))
suppressMessages(library(DT))
suppressMessages(library(highcharter))

# importing the dataset
covid_africa <- read_csv("covid_19_africa.csv")

# converting the classes of Confirmed, Deaths, and Recovered columns to class integer
covid_africa <- covid_africa %>%
        
        mutate(Region = as_factor(Region),
               Confirmed = as.integer(Confirmed), 
               Deaths = as.integer(Deaths),
               Recovered = as.integer(Recovered),
               Active = as.integer(Active))
```

## COVID-19 cases in Africa by Country since 27th of January, 2020
**Note that the lastest number of cases are those reported by end of the previous day.**
```{r covid-19 Africa}
datatable(covid_africa)
```

## Number of African countries with confirmed cases
```{r confirmed africa, echo = TRUE}
length(unique(covid_africa$Country))
```
All the 54 African countries recognized by the UN and WHO have confirmed cases of COVID-19.

**Interactive plot of the cumulative number of COVID-19 cases in Africa**
```{r cases in africa, echo = TRUE, warning=FALSE}
cases <- covid_africa %>%
        # grouping by observationdate
        group_by(ObservationDate) %>%
        
        # obtaining the sum of cases for each day
        summarise(Confirmed = sum(Confirmed), Deaths = sum(Deaths), Recovered = sum(Recovered), Active = sum(Active))

# creating a character vector of colors for the cases
colors <- c("Active" = "cyan", "Confirmed" = "orange", "Deaths" = "red", "Recovered" = "lightseagreen")

# plotting
cases %>%
        
        # placing all cases in a single column and their sums in another column 
        gather(Confirmed, Deaths, Recovered, Active, key = "cases", value = "number") %>%
        
        # setting general chart options
        hchart("line", hcaes(x = ObservationDate, y = number, group = cases),
               color = colors) %>%
        
        # setting the plot title
        hc_title(text = "Cumulative number of COVID-19 cases in Africa") %>%
        
        # setting current date as subtitle
        hc_subtitle(text = today()) %>%
        
        # setting the x axis label
        hc_xAxis(title = list(text = "Date")) %>%
                
        # setting the y axis label        
        hc_yAxis(title = list(text = "Number of cases")) %>%
        
        # add plot border
        hc_chart(borderColor = "grey",
                 borderWidth = 1) %>%
        
        # plot theme
        hc_add_theme(hc_theme_google())
```

**Interactive plot of the daily new confirmed cases and new deaths from COVID-19 in Africa.**
```{r new cases and deaths}
new_cases <- cases %>%
        mutate(new_confirmed = Confirmed - lag(Confirmed, default = first(Confirmed)),
               new_deaths = Deaths - lag(Deaths, default = first(Deaths)),
               new_recovered = Recovered - lag(Recovered, default = first(Recovered)),
               new_active = Active - lag(Active, default = first(Active))) %>%
        select(ObservationDate, new_confirmed:new_deaths)

new_colors <- c("new_confirmed" = "orange", "new_deaths" = "red")

new_cases %>%
        gather(new_confirmed, new_deaths, 
               key = "cases", value = "number") %>%
        hchart("column", hcaes(x = ObservationDate, y = number, group = cases),
               color = new_colors) %>%
        hc_chart(borderColor = "grey",
                 borderWidth = 1) %>%
        hc_xAxis(title = list(text = "Date")) %>%
        hc_yAxis(title = list(text = "Number of cases")) %>%
        hc_add_theme(hc_theme_google()) %>%
        hc_title(text = "Daily new cases and deaths from COVID-19 in Africa") %>%
        hc_subtitle(text = today())
```

**Interactive plots of the cumulative numbers of COVID-19 cases by African Region.**
```{r confirmed regional cases}
regional_cases <- covid_africa %>%
        # grouping cases by date and region
        group_by(ObservationDate, Region) %>%
        
        # obtaining sum of the cases on each date from each region
        summarise(Confirmed = sum(Confirmed), Deaths = sum(Deaths),
                  Recovered = sum(Recovered), Active = sum(Active))

# plotting
regional_cases %>%                  
hchart("line", hcaes(x = ObservationDate, y = Confirmed, group = Region)) %>%
        hc_xAxis(title = list(text = "Date")) %>%
        hc_title(text = "Cumulative number of Confirmed cases by region") %>%
        hc_chart(borderColor = "grey",
                 borderWidth = 1) %>%
        hc_subtitle(text = today()) %>%
        hc_add_theme(hc_theme_google())
```

```{r deaths regional}
regional_cases %>%                  
hchart("line", hcaes(x = ObservationDate, y = Deaths, group = Region)) %>%
        hc_xAxis(title = list(text = "Date")) %>%
        hc_title(text = "Cumulative number of COVID-19 deaths by region") %>%
        hc_subtitle(text = today()) %>%
        hc_chart(borderColor = "grey",
                 borderWidth = 1) %>%
        hc_add_theme(hc_theme_google())
```

**Interactive plots of the highest to lowest numbers of COVID-19 cases by African country.**
```{r confirmed by country, echo = TRUE, warning=FALSE}
## Latest number of cases of COVID-19 in Africa by country
latest <- covid_africa %>%
        filter(ObservationDate == today() - 1 )

latest_confirmed <- latest %>%
        
        # arranging confirmed cases in descending order
        arrange(desc(Confirmed))

# plot       
highchart() %>% 
        
        # adding series
        hc_add_series(data = latest_confirmed, type = "bar",
                      hcaes(x = Country, y = Confirmed),
                      name = "Confirmed cases", showInLegend = FALSE,
                      color = "orange") %>%
        
        # adding x axis parameters
        hc_xAxis(categories = latest_confirmed$Country,
                 title = list(text = "Country")) %>%
        
        # adding y axis parameters
        hc_yAxis(title = list(text = "Confirmed cases")) %>%
        
        # adding the plot title
        hc_title(text = "Confirmed cases of COVID-19 in Africa by Country") %>%
        
        # setting today's date as a subtitle
        hc_subtitle(text = today()) %>%
        
        # add plot border
        hc_chart(borderColor = "grey",
                 borderWidth = 1) %>%
        
        # specifying the plot size
        hc_size(700, 900)
```

```{r deaths, echo = TRUE, warning=FALSE}
latest_deaths <- latest %>%
        arrange(desc(Deaths))

# plot       
highchart() %>% 
        hc_add_series(data = latest_deaths, type = "bar",
                      hcaes(x = Country, y = Deaths),
                      name = "Deaths", showInLegend = FALSE,
                      color = "red") %>%
        hc_xAxis(categories = latest_deaths$Country,
                 title = list(text = "Country")) %>%
        hc_yAxis(title = list(text = "Deaths")) %>%
        hc_title(text = "Deaths from COVID-19 in Africa by Country") %>%
        hc_subtitle(text = today()) %>%
        hc_chart(borderColor = "grey",
                 borderWidth = 1) %>%
        hc_size(700, 900)
```

```{r recovered, echo = TRUE, warning=FALSE}
latest_recovered <- latest %>%
        arrange(desc(Recovered))

# plot       
highchart() %>% 
        hc_add_series(data = latest_recovered, type = "bar",
                      hcaes(x = Country, y = Recovered),
                      name = "Recovered cases", showInLegend = FALSE,
                      color = "lightseagreen") %>%
        hc_xAxis(categories = latest_recovered$Country,
                 title = list(text = "Country")) %>%
        hc_yAxis(title = list(text = "Recovered cases")) %>%
         hc_title(text = "Recovered cases from COVID-19 in Africa by Country") %>%
        hc_subtitle(text = today()) %>%
        hc_chart(borderColor = "grey",
                 borderWidth = 1) %>%
        hc_size(700, 900)
```

```{r active, echo = TRUE, warning=FALSE}
latest_active <- latest %>%
        arrange(desc(Active))

# plot       
highchart() %>% 
        hc_add_series(data = latest_active, type = "bar",
                      hcaes(x = Country, y = Active),
                      name = "Active cases", showInLegend = FALSE,
                      color = "cyan") %>%
        hc_xAxis(categories = latest_active$Country,
                 title = list(text = "Country")) %>%
        hc_yAxis(title = list(text = "Active cases")) %>%
        hc_title(text = "Active cases of COVID-19 in Africa by Country") %>%
        hc_subtitle(text = today()) %>%
        hc_chart(borderColor = "grey",
                 borderWidth = 1) %>%
        hc_size(700, 900)
```

## References
1. John Hopkins University Covid_19 datasets: 
https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_daily_reports
2. SRk's dataset on kaggle: https://www.kaggle.com/sudalairajkumar/novel-corona-virus-2019-dataset#covid_19_data.csv
3. African Arguments, Covid-19 Africa Tracker: 
https://africanarguments.org/2020/04/07/coronavirus-in-africa-tracker-how-many-cases-and-where-latest/







