---
title: "Interactive Visualizations of COVID-19 cases in Africa"
author: "Julius Okwir"
date: "`r format(Sys.Date())`"
output:
    html_document:
         toc: true
         toc_depth: 2
         code_folding: hide
---

# Background
Late in December 2019, the [World Health Organisation](https://www.who.int/emergencies/diseases/novel-coronavirus-2019/events-as-they-happen) (WHO) China Country Office obtained information about a severe pneumonia of an unknown cause, detected in the city of Wuhan in Hubei province, China. This later turned out to be the **novel coronavirus disease** (COVID-19), an infectious disease caused by **severe acute respiratory syndrome coronavirus-2** (SARS-CoV-2) of the corona virus family. The disease causes respiratory illness characterised by primary [symptoms](https://www.cdc.gov/coronavirus/2019-ncov/symptoms-testing/symptoms.html) like cough, fever, and in more acute cases, difficulty in breathing. **WHO** later [declared covid-19 as a Pandemic](https://time.com/5791661/who-coronavirus-pandemic-declaration/) because of its fast rate of spread across the Globe with over [5.69 Million confirmed cases and over 355,000 deaths](https://coronavirus.jhu.edu/map.html) as of May 28, 2020. The African continent started confirming its first cases in late January and mid-February of 2020 in some countries. The disease has since spread across all the 54 Africa countries with over [124,000 confirmed cases and over 3,600 deaths](https://africanarguments.org/2020/04/07/coronavirus-in-africa-tracker-how-many-cases-and-where-latest/) as of May 28, 2020. 

# The Dataset
The [covid_19_africa.csv](https://github.com/ju-ok/Covid-19-Africa.git) dataset contains daily level information about the Covid-19 cases in Africa.
It is a time series data and hence the number of cases on any given day is cumulative.I extracted the data from the [covid_19_data.csv](https://www.kaggle.com/sudalairajkumar/novel-corona-virus-2019-dataset#covid_19_data.csv) which was made available on kaggle. The R script that I used to prepare this dataset is also available on [my Github repository](https://github.com/ju-ok/Covid-19-Africa.git). The original datasets can be found on [John Hopkins University Github repository](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_daily_reports) for covid_19. 

## Field description
- ObservationDate: YY/MM/DD
- Country: name of African country
- Region: Region of the African country
- Confirmed: the number of cOVID-19 confirmed cases
- Deaths:  the number of deaths from COVID-19
- Recovered: the number of recovered cases
- Active: the number of people still infected with COVID-19

## Task
Visualizations of Africa’s COVID-19 cases.

# Data processing
```{r, message = FALSE, Warning = FALSE, message=FALSE}
# loading the required packages
suppressMessages(library(tidyverse))
suppressMessages(library(lubridate))
suppressMessages(library(DT))
suppressMessages(library(highcharter))
suppressMessages(library(htmltools))

# importing the dataset
covid_africa <- read_csv("covid_19_africa.csv")

# converting the classes of Confirmed, Deaths, and Recovered columns to class integer
covid_africa <- covid_africa %>%
        
        mutate(Region = as_factor(Region),
               Confirmed = as.integer(Confirmed), 
               Deaths = as.integer(Deaths),
               Recovered = as.integer(Recovered),
               Active = as.integer(Active))
```

```{r}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# COVID-19 in Africa
## A summary of COVID-19 cases in Africa since 27th of January, 2020
Note that the lastest number of cases are those reported by end of the previous day.
```{r covid-19 Africa}
datatable(covid_africa)
```

## Number of African countries with confirmed cases
```{r confirmed africa}
length(unique(covid_africa$Country))
```
All the 54 African countries recognized by the UN and WHO have confirmed cases of COVID-19.

## Cumulative number of COVID-19 cases in Africa
**Note: all plots with legends can be activated or deactivated by clicking on the legend.**
```{r cases in africa}
# obtaining the cumulative cases by observation date
cases <- covid_africa %>%
        group_by(ObservationDate) %>%
        summarise(Confirmed = sum(Confirmed), Deaths = sum(Deaths), 
                  Recovered = sum(Recovered), Active = sum(Active))

# specifying the plot colors
colors <- c("Active" = "cyan", "Confirmed" = "orange", "Deaths" = "red", "Recovered" = "lightseagreen")

# plotting
cases %>%
        gather(Confirmed, Deaths, Recovered, Active, key = "cases", value = "number") %>%
        hchart("line", hcaes(x = ObservationDate, y = number, group = cases),
               color = colors) %>%
        hc_title(text = "Cumulative number of COVID-19 cases in Africa") %>%
        hc_subtitle(text = today()) %>%
        hc_xAxis(title = list(text = "Date")) %>%
        hc_yAxis(title = list(text = "Number of cases")) %>%
        hc_chart(borderColor = "grey",
                 borderWidth = 1) %>%
        hc_tooltip(crosshairs = TRUE, sort = TRUE, table = TRUE) %>%
        hc_add_theme(hc_theme_google())
```

## Daily new confirmed cases and new deaths from COVID-19 in Africa
```{r new cases and deaths}
# computing the daily new cases and deaths
new_cases <- cases %>%
        mutate(new_confirmed = Confirmed - lag(Confirmed, default = first(Confirmed)),
               new_deaths = Deaths - lag(Deaths, default = first(Deaths))) %>%
        select(ObservationDate, new_confirmed, new_deaths)

# specifying the plot colors
new_colors <- c("new_confirmed" = "orange", "new_deaths" = "red")

# plotting
new_cases %>%
        gather(new_confirmed, new_deaths, 
               key = "cases", value = "number") %>%
        hchart("column", hcaes(x = ObservationDate, y = number, group = cases),
               color = new_colors) %>%
        hc_chart(borderColor = "grey",
                 borderWidth = 1) %>%
        hc_xAxis(title = list(text = "Date")) %>%
        hc_yAxis(title = list(text = "Number of cases")) %>%
        hc_add_theme(hc_theme_google()) %>%
        hc_title(text = "Daily new cases and deaths from COVID-19 in Africa") %>%
        hc_tooltip(crosshairs = TRUE, sort = TRUE, table = TRUE) %>%
        hc_subtitle(text = today())
```

## Case Fatality Rate(%) and Recovery Rate(%) from COVID-19 in Africa
The [case fatality rate](https://www.britannica.com/science/case-fatality-rate) and recovery rate are dependant on accurate reporting of COVID-19 cases. The limited testing capacities of some African countries implies that these values may be inexact.
```{r CFR and RR}
# computing the rates
rates <- cases %>%
  mutate(case_fatality = (Deaths / Confirmed) * 100,
         recovery_rate = (Recovered / Confirmed) * 100,
         CFR = round(case_fatality, 2),
         RR = round(recovery_rate, 2))

# specifying the colors
rates_color <- c("CFR" = "red", "RR" = "lightgreen")

# plotting
rates %>% 
  gather(CFR, RR, key = "cases", value = "number") %>%
  hchart("line", hcaes(x = ObservationDate, y = number, group = cases),
         color = rates_color) %>%
  hc_xAxis(title = list(text = "Date")) %>%
  hc_yAxis(title = list(text = "Percent(%)")) %>%
  hc_tooltip(crosshairs = TRUE, sort = TRUE, table = TRUE) %>%
  hc_title(text = "Changes in the Case Fatality Rate(CFR) and Recovery Rate(RR) in Africa") %>%
  hc_chart(borderColor = "grey",
                 borderWidth = 1) %>%
  hc_add_theme(hc_theme_google())
```

## Heatmaps of Africa showing COVID-19 cases {.tabset .tabset-fade .tabset-pills}
```{r heatmaps}
# getting the latest number of cases of COVID-19 in Africa by country
latest <- covid_africa %>%
        filter(ObservationDate == today() - 1 )

# downloading a map of Africa
africa_map <- download_map_data("https://code.highcharts.com/mapdata/custom/africa.js")

# looking and the African Countries
properties <- get_data_from_map(africa_map)

# modify some country names in the covid_19_africa data to match those in africa_map
latest_1 <-  latest %>%
        mutate(Country = replace(Country, Country == "Congo (Kinshasa)", "Democratic Republic of the Congo"),
               Country = replace(Country, Country == "Congo (Brazzaville)", "Republic of Congo"),
               Country = replace(Country, Country == "Tanzania", "United Republic of Tanzania"))
```
        
### Confirmed
```{r confirmed heatmap}
highchart() %>%
        hc_add_series_map(africa_map, latest_1, value = "Confirmed",
                          joinBy = c("name", "Country"), name = "Confirmed") %>%
        hc_colorAxis(minColor = "lightyellow", maxColor = "Darkorange") %>%
        hc_mapNavigation(enabled = TRUE) %>%
        hc_title(text = "Confirmed") %>%
        hc_subtitle(text = today()) %>%
        hc_chart(borderColor = "grey",
                 borderWidth = 1)
```

### Deaths
```{r deaths heatmap}
highchart() %>%
        hc_add_series_map(africa_map, latest_1, value = "Deaths",
                          joinBy = c("name", "Country"), name = "Deaths") %>%
        hc_colorAxis(minColor = "snow", maxColor = "darkred") %>%
        hc_mapNavigation(enabled = TRUE) %>%
        hc_title(text = "Deaths") %>%
        hc_subtitle(text = today()) %>%
        hc_chart(borderColor = "grey",
                 borderWidth = 1)
```

### Recovered
```{r recovered heatmap}
highchart() %>%
  hc_add_series_map(africa_map, latest_1, value = "Recovered",
                    joinBy = c("name", "Country"), name = "Recovered") %>%
  hc_colorAxis(minColor = "lightgreen", maxColor = "darkgreen") %>%
  hc_mapNavigation(enabled = TRUE) %>%
  hc_title(text = "Recovered") %>%
  hc_subtitle(text = today()) %>%
  hc_chart(borderColor = "grey",
           borderWidth = 1)
```

### Active
```{r active heatmap}
highchart() %>%
  hc_add_series_map(africa_map, latest_1, value = "Active",
                    joinBy = c("name", "Country"), name = "Active") %>%
  hc_colorAxis(minColor = "lightcyan", maxColor = "darkcyan") %>%
  hc_mapNavigation(enabled = TRUE) %>%
  hc_title(text = "Active") %>%
  hc_subtitle(text = today()) %>%
  hc_chart(borderColor = "grey",
           borderWidth = 1)
```

# COVID-19 cases by African Region
## Cumulative number of COVID-19 cases by African Region {.tabset .tabset-fade .tabset-pills}
```{r regional cases}
# computing the regional cases of COVID-19
regional_cases <- covid_africa %>%
        group_by(ObservationDate, Region) %>%
        summarise(Confirmed = sum(Confirmed), Deaths = sum(Deaths),
                  Recovered = sum(Recovered), Active = sum(Active))
```
### Confirmed
```{r confirmed by region}
# plotting
regional_cases %>%                  
hchart("line", hcaes(x = ObservationDate, y = Confirmed, group = Region)) %>%
        hc_xAxis(title = list(text = "Date")) %>%       
        hc_title(text = "Cumulative Confirmed cases") %>%
        hc_subtitle(text = today()) %>%
        hc_chart(borderColor = "grey",
           borderWidth = 1) %>%
        hc_tooltip(crosshairs = TRUE, sort = TRUE, table = TRUE) %>%
        hc_add_theme(hc_theme_google())
```

### Deaths
```{r regional deaths}
regional_cases %>%                  
        hchart("line", hcaes(x = ObservationDate, y = Deaths, group = Region)) %>%
        hc_xAxis(title = list(text = "Date")) %>%      
        hc_title(text = "Cumulative Deaths") %>%
        hc_subtitle(text = today()) %>%
        hc_chart(borderColor = "grey",
           borderWidth = 1) %>%
        hc_tooltip(crosshairs = TRUE, sort = TRUE, table = TRUE) %>%
        hc_add_theme(hc_theme_google())
```

### Recovered
```{r regional recovered}
regional_cases %>%                  
  hchart("line", hcaes(x = ObservationDate, y = Recovered, group = Region)) %>%
  hc_xAxis(title = list(text = "Date")) %>%      
  hc_title(text = "Cumulative Recoveries") %>%
  hc_subtitle(text = today()) %>%
  hc_chart(borderColor = "grey",
           borderWidth = 1) %>%
  hc_tooltip(crosshairs = TRUE, sort = TRUE, table = TRUE) %>%
  hc_add_theme(hc_theme_google())
```

### Active
```{r regional active}
regional_cases %>%                  
  hchart("line", hcaes(x = ObservationDate, y = Active, group = Region)) %>%
  hc_xAxis(title = list(text = "Date")) %>%      
  hc_title(text = "Cumulative Active cases") %>%
  hc_subtitle(text = today()) %>%
  hc_chart(borderColor = "grey",
           borderWidth = 1) %>%
  hc_tooltip(crosshairs = TRUE, sort = TRUE, table = TRUE) %>%
  hc_add_theme(hc_theme_google())
```

## Case Fatality Rate(%) and Recovery Rate(%) 
```{r regional CFR and RR}
# computing the regional CFR and RR
regional_rates <- latest %>%
  group_by(Region) %>%
  summarise(Confirmed = sum(Confirmed), Deaths = sum(Deaths), Recovered = sum(Recovered)) %>%
  mutate(case_fatality = (Deaths / Confirmed) * 100,
         fatality_rounded = round(case_fatality, 2),
         recovery_rate = (Recovered / Confirmed) * 100,
         recovery_rounded = round(recovery_rate, 2))
  
# plotting
regional_fatality <- regional_rates %>%
  arrange(case_fatality) %>%
  hchart("column", hcaes(Region, fatality_rounded), color = "red", name = "CFR(%)") %>%
  hc_yAxis(title = list(text = "Case Fatality Rate (%)")) %>%
  hc_title(text = "Case Fatality Rate by African Region") %>%
  hc_subtitle(text = today()) %>%
  hc_chart(borderColor = "grey",
           borderWidth = 1) %>%
  hc_add_theme(hc_theme_google())

regional_recovery <- regional_rates %>%
  arrange(recovery_rate) %>%
  hchart("column", hcaes(Region, recovery_rounded), color = "green", name = "RR(%)") %>%
  hc_yAxis(title = list(text = "Recovery Rate (%)")) %>%
  hc_title(text = "Recovery Rate by African Region") %>%
  hc_subtitle(text = today()) %>%
  hc_chart(borderColor = "grey",
           borderWidth = 1) %>%
  hc_add_theme(hc_theme_google())

# all plots on same page
all_rates <- list(regional_fatality, regional_recovery)
hw_grid(all_rates, rowheight = 400, ncol = 2) %>% browsable()
```

# COVID-19 cases by African Country
## Highest to the Lowest: the number of COVID-19 cases in Africa by Country {.tabset .tabset-fade .tabset-pills}
```{r latest cases}
# getting the latest number of confirmed cases in descending order
latest_confirmed <- latest %>%
        arrange(desc(Confirmed))
```
### Confirmed
```{r confirmed by country}
# plotting       
highchart() %>% 
        hc_add_series(data = latest_confirmed, type = "bar",
                      hcaes(x = Country, y = Confirmed),
                      name = "Confirmed cases", showInLegend = FALSE,
                      color = "orange") %>%
        hc_xAxis(categories = latest_confirmed$Country,
                 title = list(text = "Country")) %>%
        hc_yAxis(title = list(text = "Confirmed cases")) %>%
        hc_title(text = "COVID-19 Confirmed cases by African Country") %>%
        hc_subtitle(text = today()) %>%
        hc_chart(borderColor = "grey",
                 borderWidth = 1) %>%
        hc_size(900, 900)
```

### Deaths
```{r deaths}
# COVID-19 deaths in descending order
latest_deaths <- latest %>%
        arrange(desc(Deaths))

# plotting       
highchart() %>% 
        hc_add_series(data = latest_deaths, type = "bar",
                      hcaes(x = Country, y = Deaths),
                      name = "Deaths", showInLegend = FALSE,
                      color = "red") %>%
        hc_xAxis(categories = latest_deaths$Country,
                 title = list(text = "Country")) %>%
        hc_yAxis(title = list(text = "Deaths")) %>%
        hc_title(text = "Deaths from COVID-19 by Country") %>%
        hc_subtitle(text = today()) %>%
        hc_chart(borderColor = "grey",
                 borderWidth = 1) %>%
        hc_size(900, 900)
```

### Recovered 
```{r recovered}
# recovered cases in descending order
latest_recovered <- latest %>%
        arrange(desc(Recovered))

# plotting       
highchart() %>% 
        hc_add_series(data = latest_recovered, type = "bar",
                      hcaes(x = Country, y = Recovered),
                      name = "Recovered cases", showInLegend = FALSE,
                      color = "lightseagreen") %>%
        hc_xAxis(categories = latest_recovered$Country,
                 title = list(text = "Country")) %>%
        hc_yAxis(title = list(text = "Recovered cases")) %>%
         hc_title(text = "COVID-19 recoveries by Country") %>%
        hc_subtitle(text = today()) %>%
        hc_chart(borderColor = "grey",
                 borderWidth = 1) %>%
        hc_size(900, 900)
```

### Active
```{r active}
# active cases in ascending order
latest_active <- latest %>%
        arrange(desc(Active))

# plotting       
highchart() %>% 
        hc_add_series(data = latest_active, type = "bar",
                      hcaes(x = Country, y = Active),
                      name = "Active cases", showInLegend = FALSE,
                      color = "cyan") %>%
        hc_xAxis(categories = latest_active$Country,
                 title = list(text = "Country")) %>%
        hc_yAxis(title = list(text = "Active cases")) %>%
        hc_title(text = "Active cases of COVID-19 in Africa by Country") %>%
        hc_subtitle(text = today()) %>%
        hc_chart(borderColor = "grey",
                 borderWidth = 1) %>%
        hc_size(900, 900)
```

## Case Fatality Rate(%) and Recovery Rate(%) in Each African Country
```{r CFR and RR by country}
latest %>%
  mutate(case_fatality = (Deaths / Confirmed) * 100,
         fatality_rounded = round(case_fatality, 2),
         recovery_rate = (Recovered / Confirmed) * 100,
         recovery_rounded = round(recovery_rate, 2)) %>%
  rename("Case Fatality Rate(%)" = fatality_rounded,
         "Recovery Rate(%)" = recovery_rounded) %>%
  select(Country, Region, "Case Fatality Rate(%)", "Recovery Rate(%)") %>%
  datatable()
```

## Compare countries of Choice {.tabset .tabset-fade .tabset-pills}
### Confirmed
Compare the changes in the number of COVID-19 cases in your countries of choice
```{r compare confirmed}
covid_africa %>%
hchart("spline", hcaes(ObservationDate, Confirmed, group = Country), visible = FALSE) %>%
  hc_xAxis(title = list(text = "Date")) %>%      
  hc_title(text = "Confirmed cases") %>%
  hc_subtitle(text = today()) %>%
  hc_chart(borderColor = "grey",
           borderWidth = 1) %>%
  hc_tooltip(crosshairs = TRUE, sort = TRUE, table = TRUE) %>%
  hc_add_theme(hc_theme_google()) %>%
  hc_size(800, 600)
```

### Deaths
Compare changes in the number of COVID-19 deaths in your countries of choice
```{r compare deaths}
covid_africa %>%
  hchart("spline", hcaes(ObservationDate, Deaths, group = Country), visible = FALSE) %>%
  hc_xAxis(title = list(text = "Date")) %>%      
  hc_title(text = "Deaths") %>%
  hc_subtitle(text = today()) %>%
  hc_chart(borderColor = "grey",
           borderWidth = 1) %>%
  hc_tooltip(crosshairs = TRUE, sort = TRUE, table = TRUE) %>%
  hc_add_theme(hc_theme_google()) %>%
  hc_size(800, 600)
```

### Recovered
Compare changes in the number of COVID-19 recoveries in your countries of choice
```{r compare recoveries}
covid_africa %>%
  hchart("spline", hcaes(ObservationDate, Recovered, group = Country),
         visible = FALSE) %>%
  hc_xAxis(title = list(text = "Date")) %>%      
  hc_title(text = "Deaths") %>%
  hc_subtitle(text = today()) %>%
  hc_chart(borderColor = "grey",
           borderWidth = 1) %>%
  hc_tooltip(crosshairs = TRUE, sort = TRUE, table = TRUE) %>%
  hc_add_theme(hc_theme_google()) %>%
  hc_size(800, 600)
```

### Active
Compare changes in the number of COVID-19 Active cases in your countries of choice
```{r compare active}
covid_africa %>%
  hchart("spline", hcaes(ObservationDate, Active, group = Country),
         visible = FALSE) %>%
  hc_xAxis(title = list(text = "Date")) %>%      
  hc_title(text = "Active") %>%
  hc_subtitle(text = today()) %>%
  hc_chart(borderColor = "grey",
           borderWidth = 1) %>%
  hc_tooltip(crosshairs = TRUE, sort = TRUE, table = TRUE) %>%
  hc_add_theme(hc_theme_google()) %>%
  hc_size(800, 600)
```

# References
1. John Hopkins University Covid_19 datasets: 
https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_daily_reports
2. SRk's dataset on kaggle: https://www.kaggle.com/sudalairajkumar/novel-corona-virus-2019-dataset#covid_19_data.csv
3. African Arguments, Covid-19 Africa Tracker: 
https://africanarguments.org/2020/04/07/coronavirus-in-africa-tracker-how-many-cases-and-where-latest/







